Machine Learning for Handwriting Recognition: Algorithms and Techniques

### 2023-12-16 Manuscript/Manuscript.tex
Arrange the pages
Final Draft

\section* {Fundemental Concepts}
So what is machine learning? Basically, machine learning is a subset of the artificial intelligence that allows the systems and software to learn and improve from the dataset or experience from the user without being explicitly programmed. This learning process is driven by algorithms that would interpret and process data, extract patterns and make decisions. In handwriting recognition, the algorithms play a vital role in decoding the diversity and complexity of human handwriting. Machine learning can be divided into three main types: supervised learning, where models are trained on labeled data; unsupervised learning, which processes unlabeled data and aims to discover hidden patterns; and reinforcement learning, where models learn through trial and error.Each learning type has a unique impact on handwriting recognition.

@@

As you can see, if we first use exponential function, add them together to obtain a sum, then we use this sum as the denominator, we can get a value between 0 and 1. And this number is what we want. However, if you want these values contain the meaning of probability, or, in other words, "become the probability", you may need to add the activation function. In the example below, the code will show how the activation function works.

@@

One of the key areas of impact is the improved capability to handle the vast variability in handwriting styles. Advanced machine learning models, particularly those employing deep learning techniques like Convolutional Neural Networks (CNNs) and Recurrent Neural Networks (RNNs), are becoming increasingly adept at understanding and interpreting the nuances of individual handwriting. This means that as these models are exposed to more data, their ability to accurately recognize even the most idiosyncratic and obscure handwriting styles is greatly enhanced.

Another significant impact is the reduction in the need for extensive preprocessing of handwritten texts. Current systems often require multiple steps of preprocessing, such as denoising, normalization, and alignment, to make the text legible for recognition algorithms. However, with advancements in AI, systems are becoming more robust, capable of interpreting text with minimal preprocessing. This not only speeds up the recognition process but also expands the range of applications, enabling the processing of texts in less-than-ideal conditions, such as historical documents or quick handwritten notes.

@@

\section* {Challenges in Handwriting Recognition}
Handwriting recognition, despite its significant advancements, still faces numerous challenges that make its implementation complex and demanding. One of the primary hurdles is the inherent variability in individual handwriting styles. Unlike typed text, which follows a uniform font and spacing, handwritten text varies greatly from person to person. This variability is not just in the shape of the letters, but also in their size, slant, spacing, and the connection between characters. For instance, one individual's 'a' might resemble another's 'o', or the way one person connects letters in cursive writing might be entirely different from another's. This diversity necessitates the development of highly adaptive and sophisticated machine learning models capable of learning and recognizing a wide range of handwriting styles.

Another significant challenge is handling skewed, rotated, or poorly scanned text. In real-world applications, the input to handwriting recognition systems can come from various sources such as handwritten notes, filled forms, or historical documents, which might not always be perfectly aligned, legible, or well-scanned. Skewed or rotated text requires preprocessing steps for alignment, which can sometimes distort the handwriting, making recognition more difficult. Additionally, the presence of noise, such as smudges, paper quality, or ink blots, can further degrade the quality of the input, posing additional challenges for the recognition algorithms.

>>>>

\section* {Fundemental Concepts}
So what is machine learning? Basically, machine learning is a subset of the artificial intelligence. It allows systems and software to learn and improve from the dataset or experience from the user without being explicitly programmed. This learning process is driven by algorithms that would interpret and process data, extract patterns and make decisions. In handwriting recognition, the algorithms play a important role in decoding the diversity and complexity of human handwriting. Machine learning can be divided into three main types: supervised learning, where models are trained on labeled data; unsupervised learning, which processes unlabeled data and aims to discover hidden patterns; and reinforcement learning, where models learn through trial and error.Each learning type has a unique impact on handwriting recognition.

@@

As you can see, if we use exponential function first, add them together to obtain a sum, then we use this sum as the denominator, we can get a value between 0 and 1. And this number is what we want. However, if you want these values contain the meaning of probability, or, in other words, "become the probability", you may need to add the activation function. In the example below, the code will show how the activation function works.

@@

One of the key areas of impact is the improved capability to handle the vast variability in handwriting styles. Advanced machine learning models, especially those employing deep learning techniques like Convolutional Neural Networks (CNNs) and Recurrent Neural Networks (RNNs) we just mentioned above, are becoming increasingly adept at understanding and interpreting the nuances of individual handwriting. This means that as these models are exposed to more data, their ability to accurately recognize even the most idiosyncratic and obscure handwriting styles is greatly enhanced.\cite{Advances}

Another significant impact is the reduction in the need for extensive pre-processing of handwritten texts. Current systems often require multiple steps of pre-processing, such as denoising, normalization, and alignment, to make the text legible for recognition algorithms. However, with advancements in AI, systems are becoming more robust, capable of interpreting text with minimal preprocessing. This not only speeds up the recognition process but also expands the range of applications, enabling the processing of texts in less-than-ideal conditions, such as historical documents or quick handwritten notes.

@@

\section* {Challenges in Handwriting Recognition}
Handwriting recognition, although it has significant advancements, it still faces numerous challenges that make its implementation complex and demanding. One of the primary hurdles is the inherent variability in individual handwriting styles. Unlike typed text, which follows a uniform font and spacing, handwritten text varies greatly from person to person. \cite{Improvements}This variability is not just in the shape of the letters, but also in their size, slant, spacing, and the connection between characters. For instance, one individual's 'a' might resemble another's 'o', or the way one person connects letters in cursive writing might be entirely different from another's. This diversity necessitates the development of highly adaptive and sophisticated machine learning models capable of learning and recognizing a wide range of handwriting styles.

Another challenge is handling the skewed, rotated, or poorly scanned text. In real-world applications, the input to handwriting recognition systems can come from various sources. Such as handwritten notes, filled forms, or historical documents, which might not always be perfectly aligned, legible, or well-scanned. Skewed or rotated text requires pre-processing steps for alignment, which can sometimes distort the handwriting, making recognition more difficult. Additionally, the presence of noise, such as smudges, paper quality, or ink blots, can further degrade the quality of the input, posing additional challenges for the recognition algorithms.

\appendix
\section*{Appendix}
The custom software and algorithms developed for this article are available for review and reuse. The source code, along with detailed documentation, is hosted on a GitHub repository.

GitHub Repository Link: https://github.com/statds/course-paper-or-presentation-Bluewiseid.git

This repository includes:

Source code of the developed algorithms.

Be sure to run command "pip/pip3 install torch torchvision numpy matplotlib" before you execute the code.
